---
title: "Chapter 4 Back-End – “The Art and Strategy of Data Visualization"
author: "Jabir Ghaffar"
date: "2025-07-30"
output: html_document
---

Here’s a **rewritten Chapter 4 Back-End** that follows your style, expands on why we use `ggplot2`, explains how to choose geoms, lists common and specialized geoms, and explains the syntax in depth with a focus on **formatting and stopping common errors**.

This is much more comprehensive and instructor-focused while staying in the style you want.

---

## **Chapter 4 Back-End – “The Art and Strategy of Data Visualization”**

**Filename:** `chapter04_backend_data_visualization.Rmd`
Author: Jabir Ghaffar

---

# **Why Visualization Matters**

At this stage, your students have learned to **collect, clean, and transform data**. Now, it’s time to give that data a voice.

> **Numbers on their own are like whispers. Visualization makes them shout.**

Visualization lets you:

* Reveal **patterns** that numbers hide
* **Communicate findings quickly**
* Make data **understandable** to someone who has never coded a day in their life

---

# **Why We Use ggplot2**

### **ggplot2 = Visualization for the tidyverse**

`ggplot2` is part of the **tidyverse family**. This is important because:

* It uses the same **grammar** and logic as `dplyr`, `tidyr`, and other tidyverse tools
* It allows **pipelining (`|>`) directly into visualizations**, keeping code readable

Instead of base R’s `plot()`, `ggplot2` gives us:

* More **control over customization**
* Better **layering** (add titles, labels, themes, annotations)
* Consistent **syntax**

---

# **The Grammar of a Plot**

The key grammar steps:

```{r}
data |>
  ggplot(aes(x = variable1, y = variable2)) +
  geom_point()
```

### Breaking it down:

* **`data`** – the dataset you’re using.
* **`|>`** – pass the data into ggplot.
* **`ggplot()`** – tells R you’re creating a plot.
* **`aes()`** – aesthetic mappings: x, y, colors, size, etc.
* **`+ geom_point()`** – add a layer. Each `geom_` layer adds a visual element.

---

## **Why Use This Style?**

We always:

* Pipe down line-by-line:

```{r}
  movies |>
    ggplot(aes(x = budget, y = revenue)) +
    geom_point()
```
* This increases **readability** and **reduces mistakes**.

---

# **Choosing the Right Geom**

**Choosing a geom is like choosing a lens. Each one tells a different story:**

* **Scatterplot (`geom_point`)** – Shows relationship between two numeric variables.
* **Bar chart (`geom_col`)** – Compare categories.
* **Histogram (`geom_histogram`)** – Distribution of a single numeric variable.
* **Line chart (`geom_line`)** – Trend over time or ordered values.
* **Boxplot (`geom_boxplot`)** – Spread and outliers for categories.
* **Density plot (`geom_density`)** – Smooth curve showing distribution.
* **Area chart (`geom_area`)** – Shows cumulative data over time.
* **Violin plot (`geom_violin`)** – Combines boxplot and density in one.
* **Heatmap (`geom_tile`)** – Shows patterns across two dimensions.
* **Text labels (`geom_text`)** – Annotate plots with text.

---

# **Specialized Geoms (for later chapters)**

* `geom_smooth()` – Add regression or smoothed lines.
* `geom_jitter()` – Avoid overlapping points in scatterplots.
* `geom_bar()` – For counts only (no y variable).
* `geom_rug()` – Adds small tick marks to show distribution.
* `geom_polygon()` – Useful for shapes or maps.

---

# **Common Syntax Issues to Teach Early**

### **1. Quotation Marks**

* Use quotes when something is a **label**, not a variable.

```{r}
 movies |> 
  ggplot(aes(x = budget, y = revenue)) +
  geom_point() +
  labs( title = "Movie Budgets vs Revenue", x = "Budget ($)", y = "Revenue ($)")
```
* **No quotes** inside `aes()` unless explicitly mapping to a fixed value.

---

### **2. Plus Signs vs Pipes**

* `|>`(think of the pipe as "and then") is for **data steps** (data → ggplot).
* `+` is for **adding plot layers**.

---

### **3. Factor vs Numeric Mistakes**

* Students often forget that categorical variables (like genre) work better as factors. Use:

```{r}
movies |>
    mutate(genre_names = as.factor(genre_names))  
  
```

---

# **Examples with Explanation**

---

### **Scatterplot Example**

```{r}
movies |>
  ggplot(aes( x = budget, y = revenue)) +
  geom_point() +
  labs(title = "Movie Budget vs Revenue", x = "Budget", y = "Revenue") +
  theme_minimal()
```

* **Why scatterplot?** – Because we are comparing **two continuous numeric variables**.

---

### **Bar Chart Example**

```{r}
#your code can also be written like this, which ever one is most comfortable for you.

horror_movies |>
  ggplot(
    aes(
      x = genre_names,
      y = popularity
    )
  ) +
  geom_col() +
  labs(
    title = "Popularity by Genre",
    x = "Genre",
    y = "Popularity"
  ) +
  theme_minimal()
```

* **Why bar?** – Categories on the x-axis, numeric value on the y-axis.

---

### **Line Chart Example**

```{r}
horror_movies |>
  ggplot(
    aes(
      x = release_date,
      y = vote_average
    )
  ) +
  geom_line() +
  labs(
    title = "Average Rating Over Time",
    x = "Release Date",
    y = "Vote Average"
  ) +
  theme_minimal()
```

* **Why line?** – To show change over time.

---

# **Plotting is Storytelling**

The choice of **geom** and **labels** should be guided by:

* What story are you telling?
* What insight do you want others to see first?

---

# **DP Distribution for Chapter 4**

* Mini-quests: 60 DP
* Mini-boss: 40 DP
* Total: 100 DP

---

**Next Chapter Preview:**
Chapter 5 – *Telling the Whole Story with Data*: Combining import, cleaning, transformation, visualization, and communication into one project.
