---
title: "Chapter 02: The Lens of Truth – Cleaning & Preparing Data"
author: "Jabir"
date: "2025-07-28"
output: html_document
---

# 🎯 Chapter 2 – Cleaning & Preparing Data

Welcome, Data Apprentice!
You’ve learned how to look at data, but now comes a new trial: **How do you focus when your dataset has *so many columns*?**

This chapter is all about **choosing what matters, filtering, and preparing data for your first real insights.**

------------------------------------------------------------------------

# 🗡️ Stage 1: Equipping the `tidyverse`

Our toolkit for this chapter is the **tidyverse**, which includes packages like `dplyr`, `tidyr`, `ggplot2`, and more. Think of these as your weapons and armor – they will make data wrangling readable and efficient.

Get data from Jabir Ghaffar Github: [JabirG](https://github.com/jabirG/Tidy-Tuesday-Horror-Movies-Dataset)

```{r}
# Load the tidyverse and read in your data



```

------------------------------------------------------------------------

# 🧼 Stage 2: Looking at the Dataset

Our horror movies dataset has already been cleaned for you. Now, let’s inspect it:

```{r}
# Peek at the first few rows

```

You’ll see columns like `title`, `release_date`, `popularity`, `revenue`, `runtime`, and more.

------------------------------------------------------------------------

# 🧠 Stage 3: Selecting Columns (Focus)

Data often comes with too much information. To make progress, you must **choose the few pieces of information that matter most for your story**.

In R, this is where `select()` comes in.

```{r}
# Select only the columns you want to focus on


```

This narrows down your dataset into **the essentials**.

------------------------------------------------------------------------

# ⏳ Stage 4: Filtering Data

Let’s focus only on movies **released after the year 2000**. To do this, we’ll use the **lubridate package** (a date/time toolkit).

```{r}


```

------------------------------------------------------------------------

# 📊 Stage 5: Sorting Data

You can now order your data using `arrange()` to see which movies stand out the most.

```{r}



```

------------------------------------------------------------------------

# 🧙‍♂️ MINI-BOSS: **The Lens of Truth**

> **Boss Backstory** The Lens of Truth doesn’t make your data dirty; it floods you with so much information that you cannot see what is important.
>
> To defeat this boss, you must sharpen your focus.

### **Mission Objectives**

1.  Use `lubridate` to work with the `release_date` column.
2.  Select **5–7 columns that matter most**.
3.  Filter for movies released **after the year 2000**.
4.  Arrange your data by either **popularity** or **revenue**.
5.  Write a **short explanation** (as a comment) about why you chose these columns.

```{r}
# Your code goes here:
# Step 1: Load lubridate (already done above)
# Step 2: Select your chosen columns
# Step 3: Filter for movies released after 2000
# Step 4: Arrange by popularity OR revenue



# Why these columns? Explain below as a comment:
# ...
```

------------------------------------------------------------------------

### **Completion Checklist (0–100%)**

-   [ ] Use `lubridate` (20%)
-   [ ] Select 5–7 key columns (20%)
-   [ ] Filter movies after 2000 (20%)
-   [ ] Sort by a chosen metric (20%)
-   [ ] Explain why your chosen columns matter (20%)

**HP:** 100 Each checklist item deals 20 HP. Reduce HP to 0 to defeat the boss.

**Reward:** 40 Data Points **Rank:** Apprentice → **Data Explorer**

------------------------------------------------------------------------

# 🧾 Recap

-   You learned to focus on important columns with `select()`.
-   You filtered rows with `filter()` and `lubridate`.
-   You sorted your dataset with `arrange()`.

These skills together help you transform messy datasets into something meaningful.

------------------------------------------------------------------------

# 🔮 Preview: Backend Chapter

In the **Backend Chapter for Chapter 2**, we’ll explain **why these functions exist**, the philosophy behind the tidyverse, and how to find help when you’re stuck.

------------------------------------------------------------------------

# ⏭️ Coming Up Next: Chapter 3 – Data Transformation & Exploration

The next chapter takes your clean dataset and helps you **transform and explore** it: grouping, summarizing, and discovering patterns.

------------------------------------------------------------------------
