---
title: "Side Quest Project: Style & Precision – Advanced Plot Customization"
author: "Jabir"
date: "2025-07-28"
output: html_document
---

# 🎨 Side Quest Project – Style & Precision
*(Extension of Chapter 4 – Visualization & Storytelling)*

**Goal:**  
Make your plots **more beautiful, clear, and inclusive**.  
This quest will also introduce **Posit Cloud customization** to help you personalize your coding experience.

> Complete all stages to earn **100 DP (Advanced)**.

---

# Stage 0 (Optional) – Personalize Posit Cloud (0 DP)

Your environment matters!  
Follow these steps:

1. At the top of Posit Cloud, click **Tools**.
2. Click **Global Options**.
3. Select **Appearance**.
4. Choose a theme or color scheme that makes coding enjoyable.

This won’t give you DP, but it can make your coding journey more comfortable.

---

# Stage 1 – Base Plot (15 DP)

We’ll use the **`mtcars`** dataset (built-in R dataset) to create a simple bar chart of the number of cylinders in cars.

```{r}
library(ggplot2)

mtcars |>
  ggplot(aes(x = factor(cyl))) |>
  geom_bar()
````

---

# Stage 2 – Color Customization (25 DP)

Now, let’s customize the **bar colors**.

```{r}
mtcars |>
  ggplot(aes(x = factor(cyl))) |>
  geom_bar(fill = "#69b3a2") |>  # soothing teal
  labs(
    title = "Car Count by Cylinders",
    x = "Cylinders",
    y = "Count"
  )
```

---

### Your Task:

Choose **two custom colors** (one for bars, one for the background).
Update the code above with your own color scheme.

```{r}
# Your code here

```

> **Earn 25 DP:** Custom plot with your chosen color scheme.

---

# Stage 3 – Accessibility for Color Blind Audiences (20 DP)

Use a **color-blind friendly palette**.

Install and load **`viridis`**:

```{r}
install.packages("viridis")  # only if you haven't already
library(viridis)

mtcars |>
  ggplot(aes(x = factor(cyl), fill = factor(cyl))) |>
  geom_bar() |>
  scale_fill_viridis(discrete = TRUE) |>
  labs(
    title = "Accessible Colors",
    x = "Cylinders",
    y = "Count",
    fill = "Cylinders"
  )
```

**Your Task:**
Use the `viridis` palette but choose **option = "plasma"`or`"magma"\` instead of default**.

```{r}
# Your code here

```

> **Earn 20 DP:** Correctly use an accessible palette with a new option.

---

# Stage 4 – Adding Interactivity (30 DP)

Make your plot **interactive** with `plotly`.

```{r}
install.packages("plotly")  # run once
library(plotly)

p <- mtcars |>
  ggplot(aes(x = factor(cyl), fill = factor(cyl))) |>
  geom_bar() |>
  scale_fill_viridis(discrete = TRUE)

ggplotly(p)
```

Now when you hover, you’ll see tooltips.

---

### Your Task:

Make an **interactive plot of mpg vs hp** with points colored by `cyl`.

```{r}
# Your code here

```

> **Earn 30 DP:** If your interactive plot works and shows correct information.

---

# Stage 5 – Mini Boss: **The Designer** (10 DP)

> *"I love your data… but can you make it **look like art**?
> Give me one stunning plot that uses a clean theme, custom colors, and labels that make sense. Surprise me!"*

**Instructions:**

1. Choose **any dataset** you want.
2. Create a **plot of your choice**.
3. Use:

   * Custom colors
   * A clean theme (`theme_minimal()`, `theme_light()`, etc.)
   * Labels that explain what’s shown
4. Add **one interactive element** (if possible).

---

# 💰 Total Data Points

* Stage 1: 15 DP
* Stage 2: 25 DP
* Stage 3: 20 DP
* Stage 4: 30 DP
* Mini Boss: 10 DP

**Total: 100 DP**

---

# Reflection

This quest showed you how to:

* Make your plots visually appealing
* Support accessibility with color-blind palettes
* Add interactivity for presentations
* Create visualizations that tell a story

Good design can be the difference between a **boring chart** and one that **wins a data battle!**


## Key Concepts Taught in This Quest:

- **Plot customization with fill/colors**
- **Choosing colors for accessibility (viridis)**
- **Making plots interactive (plotly)**
- **Personalizing coding environment (optional)**

